---
title: "Tidy Data"
author: "JJB + Course"
date: "10/03/2018"
output:
   html_document:
     toc: true
     toc_float:
       collapse: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pipe Operator

## Example: Piping Operator

```{r example-funcs}
# install.packages("magrittr")
library("magrittr")
4 %>%           # Take the number four and, then
    sqrt()      # find the square root

# Same as
# sqrt(4)

c(7, 42,  1,  25) %>%   # Combine four elements and, then
    log() %>%           # take the natural log and, then
    round(2) %>%        # round to the second decimal and, then
    diff()              # take the difference between consecutive elements 

# Same as
# diff(round(log(c(7,42,1,25)), 2))
```

## Example: Multi-step forms

```{r multi-step-funcs}
# Embedded / Nested Functions
set.seed(821)
mean(rnorm(10))


# Temporary Intermediate Variables
set.seed(821)
rand_nums = rnorm(10)
mean_nums = mean(rand_nums)

# Piped
# install.packages("magrittr")
library("magrittr")
set.seed(821)
rnorm(10) %>%   # Generate 10 random values from a normal and, then 
    mean()              # take the mean.
```

## Example: Inside-out form of Pipe

**Embedded/Nested** Function Calls

```{r nested-funcs, eval = FALSE}
pickup(                          # Step 5
  goto(                          # Step 4
    order(                       # Step 3
      store(                     # Step 2
        drink("Java Chip Frap"), # Step 1
        loc = "Green St.")
      )                                          
    )                                            
  )
```


**Piped**

```{r piped-funcs, eval = FALSE}
"Java Chip Frap" %>% 
  drink() %>%                  # Step 1
  store(loc = "Green St.") %>% # Step 2
  order() %>%                  # Step 3
  goto() %>%                   # Step 4
  pickup()                     # Step 5
```


## Example: Pipe to Other Arugments

```{r pipe-other-args}
subtract_vals = function(x, y) {
   x - y
}

x = 7; y = 4
subtract_vals(x, y)

x %>% subtract_vals(y)   # Default, e.g. subtract_vals(x, y)

x %>% subtract_vals( . , y) # Default, e.g. subtract_vals(x, y)
                   # ^ Period specifies where the value should go

y %>% subtract_vals(x, . ) # Pipe y to second argument, e.g. subtract_vals(x, y)
                     # ^ Period specifies where the value should go

x %>% subtract_vals(y, . ) # Pipe x to second argument, e.g. subtract_vals(y, x)
                     # ^ Period specifies where the value should go
```

## Example: Accessing Values in Data Mid-Pipe

```{r pipe-data-access}
my_df = data.frame(x = c(0, 1), y = c(2, 3))

my_df %>% .[["x"]]
        # ^ Period specifies where the data frame should go

# Equivalent to
my_df$x
my_df[["x"]]
my_df[[1]]

# Nested period usage to retrieve the last column
my_df %>% .[[ncol(.)]]
        # ^       ^ Period specifies where the data frame should go

# Equivalent to
my_df[[ncol(my_df)]]
```


### Exercise: Working with Pipes

Make the following pipeable

```{r}
# install.packages("dplyr")
library("dplyr")

# How can we transform this embedded code statement?
tail(filter(iris, Petal.Width > mean(Petal.Width)))

# We want to transform it using %>%

tail(
  filter(
    iris, Petal.Width > mean(Petal.Width)
    )
  )

# Take the inside result and switch it with the outside portion
# a.k.a inverting the function order

iris %>%
  filter(Petal.Width > mean(Petal.Width)) %>%
  tail()


```

Write a pipe that provides the `sqrt` of `2+2`

```{r}
sqrt(2 + 2)

2 + 2 %>%
  sqrt()

2 + sqrt(2)


(2 + 2) %>%
  sqrt()
```


## Example: Enrollment Untidy to Tidy

```{r transform-enrollment}
# install.packages("tidyr")
library("tidyr")

# Untidy Data
enrolled_fa17 = data.frame(
  undergrads = c(18345, 15267, 12),
  profs = c(352, 640, 0),
  grads = c(7173, 6028, 9),
  gender = c("Men", "Women", "Unknown")
)

# Why do we need to include gender inside the data.frame? 

# Tidy the data
enrolled_fa17_tidy = gather(enrolled_fa17,
                              key = "Year",       # What the key is
                              value = "Enrolled", # Specify what the _third variable_ should be
                              undergrads:grads)   # Take variables from undergrads to grads
                                                  # similar to seq(from, to)

enrolled_fa17_tidy
```

## Example: Alternative Meanings - ChickWeights

```{r transform-chicks}
# install.packages("tidyr")
library("tidyr")

# "Long"-form or "Tidy Data"
head(ChickWeight)

# "Wide"-form or "Messy data"
ChickWeight_wide = ChickWeight %>% 
  spread(Time, weight)

# Check data
head(ChickWeight_wide)

# Recover "long"-form or "Tidy Data"
ChickWeight_long = ChickWeight_wide %>% 
  gather(key = Time,      # Key for the key/value pairing
         value = weight,  # Column for measurements
         `0`:`21`)        # Column seleciton


# Check data
head(ChickWeight_long)
```

## Example: Alternative Meanings - Science!

Load in "Wide Data"

```{r load-wide-data}
experiment = read.table(header=TRUE, text='
 subject sex control a b
       S1   F     4.2  4.1  2.2
       S2   M     5.9  7.2  6.8
       S3   M     9.1  9.8  10.2
       S5   F     2.1  23.5  5.2
')

# Show wide-experiment data
experiment
```

Convert experiment data to long format

```{r wide-to-long-exp}
experiment_long = gather(experiment, condition, measurement, control:b)

# Example of Wide Format
head(experiment_long)
```


### Exercise: Making a Data Set Messy

Make the `mtcars` data set messy by converting it to:

```
#        model type   value
# 1 AMC Javelin  mpg  15.200
# 2 AMC Javelin  cyl   8.000
# 3 AMC Javelin disp 304.000
# 4 AMC Javelin   hp 150.000
# 5 AMC Javelin drat   3.150
# 6 AMC Javelin   wt   3.435
```

```{r convert-mtcars-data-to-long, eval = FALSE}
library("tidyr")

rownames(mtcars)
# Move the rowname to a variable name inside the data set.
mtcars$model = rownames(mtcars)
head(mtcars)

mtcars_long = mtcars %>%
  gather(key = "____", value = "____", ____:____)

# In IDE data viewer
# View(mtcars_long)
```

Now, fix the data by converting it back.

```{r convert-mtcars-data-to-wide, eval = FALSE}
mtcars_wide = mtcars_long %>%
  spread(key = "_____", value = "____")

head(mtcars_wide)
```


## Example: Splitting Values - Location Data

```{r breakdown_loc}
cities = data.frame(stringsAsFactors=FALSE,
         city = c("Houston", "Miami", "Atlanta", "Chicago", "Los Angeles",
                  "Washington, D.C.", "New York"),
          loc = c("29.81997438, -95.33997929", "25.7876107, -80.22410608",
                  "33.83001385, -84.39994938", "41.82999066, -87.75005497",
                  "33.98997825, -118.1799805", "38.89954938, -77.00941858", 
                  "40.74997906, -73.98001693"),
          pop = c(4053287, 2983947, 2464454, 5915976, 8097410, 2445216.5,
                  13524139),
         iso3 = c("USA", "USA", "USA", "USA", "USA", "USA", "USA"),
     province = c("Texas", "Florida", "Georgia", "Illinois", "California",
                  "District of Columbia", "New York")
)

cities_split = cities %>% 
  separate(loc, c("lat", "lng"), sep=",")

cities_split
```

## Example: Uniting Values - Location Data

```{r combine_locs, dependson = "breakdown_loc"}
cities_split %>% unite(loc, c("lat", "lng"), sep = ",")
```


### Exercise: Tidying WHO data

```{r who-messy, eval = FALSE}
who = tidyr::who

View(who)
head(who)
library("ggplot2")
library("stringr")

colnames(who)

# Convert from a combined value to a split value.
colnames(who) = str_replace_all(colnames(who), pattern = "newrel", replacement = "new_rel")

# Check to see that files are all in alignment e.g. new_sp_mXXXX
colnames(who)

# Matching codes
# newrel => new_rel

tidied_who = who %>% 
  gather(key = "____",    # Key is where we should fold values under 
         value = "____",  # Value is where the observation is
         ____:____,       # Specifying the range of column names to transform
         na.rm = TRUE) %>% # We remove any case that has a missing value
  separate(key, c("____", "____", "____")) %>% # Split apart variables
  separate(sexage, c("____", "____"), sep = 1) # Break apart more variables
```



