---
title: "Grammar of Data"
author: "JJB + Course"
date: "10/10/2018"
output:
   html_document:
     toc: true
     toc_float:
       collapse: false
---

# dplyr

```{r}
# install.packages("dplyr")
library("dplyr")
```

## Example: Enrollment Data

```{r view-enrollment}
library("tidyr")

# Untidy Data
enrolled_fa17_untidy = data.frame(
  undergrads = c(18345, 15267, 12),
  profs = c(352, 640, 0),
  grads = c(7173, 6028, 9),
  Gender = c("Men", "Women", "Unknown")
)

# Why do we need to include gender inside the data.frame? 

# Tidy the data
enrolled_fa2017 = enrolled_fa17_untidy %>% 
  gather(key = "Year",       # What the key is
         value = "Enrolled", # Specify what the _third variable_ should be
         undergrads:grads)   # Take variables from undergrads to grads
                             # similar to seq(from, to)

enrolled_fa2017


enrolled_fa2017[, 3]

enrolled_fa2017[, c(FALSE, FALSE, TRUE)]

enrolled_fa2017[, "Enrolled"]

enrolled_fa2017$Year == "undergrads"

enrolled_fa2017[enrolled_fa2017$Year == "undergrads", ]

enrolled_fa2017[enrolled_fa2017$Enrolled == max(enrolled_fa2017$Enrolled), ]

enrolled_fa2017
```


## Example: NSE

```{r nse-magic}
x = 4
x + 3

# Capture the R expression
quote({ x + 3 })

# Store the R expression for later use
my_r_command = quote({ x + 3 })

# Retrieve the R expression
my_r_command
```


## Example: Filter Observations

```{r filter-obs}
enrolled_fa2017_men = filter(enrolled_fa2017, Gender == "Men") 
enrolled_fa2017_men

enrolled_fa2017_women = filter(enrolled_fa2017, Gender == "Women")           # dplyr
enrolled_fa2017_women = enrolled_fa2017[enrolled_fa2017$Gender == "Women", ] # base R
enrolled_fa2017_women
```

## Example: Select Variables

```{r select-vars}
enrolled_fa2017_women_info = select(enrolled_fa2017_women, Year, Enrolled)
enrolled_fa2017_women_info

enrolled_fa2017_women[, c("Year", "Enrolled")]  # base R
```

## Example: Arranging Data

```{r ordered-data}
# Ascending order (lowest to highest)
enrolled_fa2017_women_ordered = arrange(enrolled_fa2017_women_info, Enrolled)
enrolled_fa2017_women_ordered

enrolled_fa2017_women[order(enrolled_fa2017_women$Enrolled), ]  # base R


# Descending order (highest to lowest)
enrolled_fa2017_women_ordered = arrange(enrolled_fa2017_women_info, desc(Enrolled))
enrolled_fa2017_women_ordered
enrolled_fa2017_women[order(enrolled_fa2017_women$Enrolled, decreasing = TRUE), ]  # base R

```

## Example: Summarising Data

```{r summarized-data}
enrolled_fa2017_total_women = summarise(enrolled_fa2017_women_info,
                                        Total_Enrolled = sum(Enrolled),
                                        Avg_Enrolled = mean(Enrolled),
                                        SD_Enrolled = sd(Enrolled))

enrolled_fa2017_total_women


sum(enrolled_fa2017_women_info$Enrolled) # base R
```


## Example: Mutating Data

```{r mutated-data}
enrolled_fa2017_women_add = mutate(enrolled_fa2017_women_info,
                                   Additional = Enrolled + 550)
enrolled_fa2017_women_add


# Directly overwrite a variable
mutate(enrolled_fa2017_women_info,
       Enrolled = Enrolled + 550)
enrolled_fa2017_women_info

enrolled_fa2017_women_add$Additional + 550 # base R
```

## Example: dplyr with pipes!

```{r full-chain}
enrolled_fa2017_total_women = 
     enrolled_fa2017 %>%                      # Take the enrollment data and, then
    filter(Gender == "Women") %>%             # Retrieve all Women data and, then
    select(Year, Enrolled) %>%                # Take Year and Enrolled variables and, then
    arrange(Enrolled) %>%                     # Order Enrolled in Ascending order and, then
    summarise(Total_Enrolled = sum(Enrolled)) # Get total women enrollment

enrolled_fa2017_total_women
```

### Exercise:

Select the Sepal.Length and Petal.Length variables in the iris data set

```{r select-variables}
head(iris)
```

```{r}
iris %>% 
  select(Sepal.Length, Petal.Length)
```

```{r}
iris[, c("Sepal.Length", "Petal.Length")]
```



Retrieve all of the `virginica` Species observations from `iris`

```{r view-iris}
head(iris)

iris %>%
  filter(Species == "virginica")
```

### Exercise: Split-Apply-Combine

Provide the mean, maximum, minimum of the Sepal.Length for each of species of iris alongside a count.

```{r split-data}
library("dplyr")
head(iris)

# Switch it to a tibble to track properties
(iris_tbl = as_tibble(iris))

iris_tbl %>%
  group_by(Species) %>%
  summarise(mean_sepal = mean(Sepal.Length), 
            max_sepal = max(Sepal.Length),
            min_sepal = min(Sepal.Length),
            n_count = n(),
                    # ^^^ Count by group observations
            n_species = n_distinct(Species)
            )
# Do not use count within a summarise
iris_tbl %>%
  group_by(Species) %>%
  summarise(n_count = count(Species))

# Count replicate the table outside of it
iris_tbl %>%
  count(Species)
```

### Example: Selection


```{r}
iris_tbl %>%
  select(-Sepal.Length, -Species)
```

```{r}
iris_tbl

iris_tbl %>%
  select(Sepal.Length, Sepal.Width:Petal.Width)
```

```{r}
iris_tbl

iris_tbl %>%
  select(1, 2)
```



