---
title: "Joins"
author: "JJB + Course"
date: "10/12/2018"
output:
   html_document:
     toc: true
     toc_float:
       collapse: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Split Data

## Example: Student Data

```{r}
Students = data.frame(id        = 1:3,
                      firstname = c("Billy", "Theodore", "Keeya"),
                      lastname  = c("Joe", "Squirrel", "Nod"),
                      age       = c(23, 25, 21),
                      instate   = c(FALSE, TRUE, TRUE))
Students
```

```{r}
Grades = data.frame(student_id = c(1, 2, 1, 3),
                    course_id  = c("STAT385", "STAT 432", "HIST100", "STAT385"),
                    grade      = c("A+", "A-", "A", "B+"))

Grades
```

```{r}
library("dplyr")


dplyr::left_join(Students,
                 Grades, 
                 by = c("id" = "student_id") )
```


```{r}
# Same number of rows, exact ordering, no repeated columns.
first_df = data.frame(A = c(1, 2, 3, 4), 
                      B = c("A", "B", "C", "A"))
first_df
```
```{r}
sec_df = data.frame(D = c(38.4, 39.9, 40, 20.5))
sec_df
```

```{r}
# Merge the data together
merged_df = data.frame(first_df, sec_df)
merged_df
```


```{r}
# Merge the data together
merged_df = data.frame(sec_df, first_df)
merged_df
```


```{r}
# Or, bind by column
merged_df_cols = cbind(first_df, sec_df)
merged_df_cols
```

```{r}
# Retrieve specific columns with the same order
selected_df = data.frame(first_df$A, sec_df$D)

selected_df
```


```{r}
# Same number of rows, exact ordering, no repeated columns.
bad_first_df = data.frame(A = c(4, 3, 2, 1),
 B = c("A", "C", "B", "A"))
bad_first_df
```

```{r}
bad_sec_df = data.frame(A = c(2, 1, 4, 3),
 D = c(39.9, 38.4, 20.5, 40))
bad_sec_df
```

```{r}
data.frame(bad_first_df,bad_sec_df)
```


```{r}
# Order data frames
ordered_first_df = bad_first_df[order(bad_first_df$A), ]
ordered_sec_df = bad_sec_df[order(bad_sec_df$A), ]
# Combine the ordered data frames
ordered_merged_df = data.frame(ordered_first_df$A,
 ordered_first_df$B,
 ordered_sec_df$D)
```


## Example: Point/Grade Data

```{r}
X = data.frame(Key = 1:3, 
               Points = c(90, 84, 75))

Y = data.frame(Key = c(1,2,4), 
               Letters = c("A", "B", "D"))
```

# Mutating Joins

## Example: Inner Join

```{r}
dplyr::inner_join(X, Y, by = "Key")
merge(X, Y, by = "Key")
```

## Example: Full

```{r}
dplyr::full_join(X, Y, by = "Key")
merge(X, Y, by = "Key", all.x = TRUE, all.y = TRUE)
```

## Example: Left

```{r}
dplyr::left_join(X, Y, by = "Key")
merge(X, Y, by = "Key", all.x = TRUE)
```

## Example: Right

```{r}
dplyr::right_join(X, Y, by = "Key")
merge(X, Y, by = "Key", all.y = TRUE)
```

## Example: Equivalence

Note the reordering of the columns...

```{r}
dplyr::left_join(X, Y, by = "Key")
```

```{r}
dplyr::right_join(Y, X, by = "Key")
```

# Example: Filtering Joins

```{r}
dplyr::anti_join(X, Y, by = "Key")
```


```{r}
dplyr::semi_join(X, Y, by = "Key")
```



# `dplyr`, `dbplyr` and RDBMS systems 


## Example: Direct Querying

Within code chunk, we're setting up our database. 

```{r setup-db}
# install.packages(c("dplyr", "dbplyr", "RSQLite"))
library("RSQLite")
library("dplyr")

# Setup database and use local data sets
con = DBI::dbConnect(RSQLite::SQLite(), path = ":dbname:")

# Copy to database via dplyr
dplyr::copy_to(con, mtcars, "mtcars")
dplyr::copy_to(con, iris, "iris")

# View listed tables
dbListTables(con)
```

To work with a SQL database and `dplyr` we need to create a source via 
`tbl`. 

```{r}
# Create a table object for dplyr
db_mtcars = tbl(con, "mtcars")

# View table
db_mtcars

# Equivalent to: 
# dbGetQuery(con, 'SELECT * FROM mtcars')
```

We can also directly embed SQL into an R Markdown document by specifying a connection...

```{sql, connection=con}
SELECT * FROM mtcars LIMIT 5
```

In _SQL_-land, there is a preference to SPEAK IN CAPITAL LETTERS. However, _SQL_ is largely case insensitive. Thus, the same query in lower letters will also work

```{sql, connection=con}
select * from mtcars limit 5
```

```{sql, connection=con}
SELECT mpg, disp FROM mtcars ORDER BY mpg
```

## Example: Lahman Data

The [Lahman data set](http://www.seanlahman.com/baseball-archive/statistics/) is compiled yearly into a SQLite-variant.

https://github.com/jknecht/baseball-archive-sqlite

```{r work-with-lehman}
# Download the data into your project workspace
# download.file("https://github.com/jknecht/baseball-archive-sqlite/blob/master/lahman2016.sqlite?raw=true", "lahman2016.sqlite")

library("dplyr")
db_lahman = DBI::dbConnect(RSQLite::SQLite(), dbname="lahman2016.sqlite")

# View listed tables
dbListTables(db_lahman)

# Specify a table
db_batting = tbl(db_lahman, "Batting")

# Compute averages
db_batting %>%
  summarise(avg_hr = mean(HR),
            avg_h = mean(H))

```

There is an underlying query process that can be shown.

```{r view-query}
# Retrieve specific values
top100 = db_batting %>%
  head(100) %>%
  select(playerID, yearID, AB, H, HR)

# Show data
top100

# View underlying query
top100 %>% 
  show_query()
```




## Example: More Direct Access via `DBI`

```{r}
library(DBI)

# Create a SQLite data base in memory 
mydb = dbConnect(RSQLite::SQLite(), path = ":memory:")

# Copy data.frame's to the database
dbWriteTable(mydb, "X", X)
dbWriteTable(mydb, "Y", Y)

# View tables
dbListTables(mydb)

# Query database
dbGetQuery(mydb, 'SELECT * FROM X WHERE "Points" < :x', 
  params = list(x = 90))

dbGetQuery(mydb, 'SELECT * FROM X WHERE "Points" < :x', 
  params = list(x = 90))
```


