---
title: 'hw02: Data Sphinx!'
author: 'STAT 385, Fall 2018'
date: 'Due: Friday, September 14th, 2018 at 6:00 PM'
output:
  html_document:
    theme: readable
    toc: yes
    toc_float:
      collapse: false
always_allow_html: yes # required for PDF
---

```{r setup, include=FALSE}
# Force all code to be displayed
knitr::opts_chunk$set(echo = TRUE)

# Enable inline testing
library(testthat)
options(testrmd_log = TRUE)
testrmd::init()
```

# Overview 

Please see the [homework policy](http://stat385.thecoatlessprofessor.com/homework-policy/)
for detailed instructions and some grading notes. Failure to follow instructions
will result in point reductions. In particular, make sure to commit each 
exercise as you complete them. 

> "Talk is cheap. Show me the code."
>
> --- **Linus Torvalds**, [on lkml](https://lkml.org/lkml/2000/8/25/132) August 25th, 2000

## Objectives 

The objectives behind this homework assignment are as follows:

- Reading in Data;
- Vectorization;
- Recycling;
- Writing Functions;
- Using R Packages

## Grading

The rubric CAs will use to grade this assignment is:

| Task                                                   | pts |
|:-------------------------------------------------------|----:|
| Link to GitHub Repository   	                         | 2   |
| At least one commit per exercise (more is better!)     | 5   |
| Commit messages that describe what changed	           | 5   |
| I _vant_ to read your data                             | 10  |
| The Recycler                                           | 17  |
| Scientists Rock!                                       | 8   |
| Winter is Coming                                       | 10  |
| Excellency at UIUC                                     | 14  |
| Total                                                  | 71  |

## Note on Markdown

If you need help with markdown syntax, please go to the "Help" menu and select the 
_Markdown Quick Reference_ guide. This will open in the **Help** tab on
the _lower-right_ corner of _RStudio_. For more examples, please see 
[the literate programming slides](http://stat385.thecoatlessprofessor.com/lectures/02-literate-programming/02-literate-programming.pdf#page=26) and the  [in class examples of writing in _RMarkdown_](https://dl.dropboxusercontent.com/s/fb36l0isvhoko8e/08-31-2018-rmarkdown-sample.Rmd?dl=0).

In addition, the following two RStudio Cheatsheets will be helpful:

- [R Markdown Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf)
- [R Markdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

Both of these guides will be given to you during your time in the
CBTF.

## Package usage

For this homework assignment, you may only use the following _R_ packages:

```{r}
pkg_list = c("ggplot2", "rmarkdown", "readxl", "nasaweather")
```

-----

# Assignment

### Collaborators

If you worked with any other student in preparing these answers, please
make sure to list their full names and NetIDs (e.g. `FirstName LastName (NetID)` ).

## (12 Points) Exercise 0: Knowledge is Power

- **[2 Points] (a)** Place a link to your `hw02` GitHub repository here.


- **[5 Points] (b)** Commit every exercise as you finish them. There are
    _five_ exercises (including this one).

- **[5 Points] (c)** Make each commit message _meaningful_. 
    - The bare minimum for a "meaningful" commit is a length of 15 characters.
    - Inside the commit message, please make sure to appropriately describe 
      what is happening.
          - Simply stating, "Exercise 3" or "Ex3" is not sufficient.
    - Provide detail on what _changed_.

## (10 Points) Exercise 1: I _vant_ to read your data! 

**[5 Points] (a)** The Division of Management Information (DMI) is trying
to use _R_ to understand the Freshman Profile of Fall 2018. They are having
difficult time importing the data into _R_ and creating a visualization.
Help them out by:

- **[3 Points]** translate the data given below into an _R_ `data.frame` called `fa17_enrolled`;
- **[1 Point]** cleaning up the visualization code; and
- **[1 Point]** allow the visualization code chunk to execute.

_Hint_: Is `ggplot()` part of base R or is it an extension package to _R_?

```bash
Year  NEnrolled 
1978	1
1990	1
1991	1
1993	2
1995	1
1996	7
1997	102
1998	2496
1999	4821
2000	84
2001	2
```


```{r my-vis, eval = FALSE}
g = ggplot(<data-set-here>,
       aes(x = Year, y = NEnrolled)) +
  geom_col()

g
```

```{r import-enrolled-data-check, test = TRUE}
target_df = readRDS("test/fa2018-enrolled.rds")
expect_true(exists("g"), info = "Verify graph was created.")
expect_equal(fa17_enrolled, target_df, info = "Verify data frame was created correctly")
```


**[5 Points] (b)** One of your international collaborators wanted to use 
the UIUC international student data at their school. However, the excel 
spreadsheet that was sent over contains multiple sheets. Your collaborator is
only interested in the Spring 2017 enrollment information. 

- **[2 Points]** Help your colleague out by reading into _R_ only the `Spring2017` sheet from `enroll_global.xlsx`. Call the imported data `sp17_enrolled`.
- **[1 Point]** Provide your colleague with a summary of the imported data. Save the
  summary into `overview_sp17`.
- **[2 Points]**Which variable (outside of "All") had the _largest_ overall amount of students?
  How about the _smallest_ **mean** number of students? (Outside of 0).




```{r enrolled-sp17}
# your code here
```

```{r check-info, test = TRUE}
# Load secured data
target_df = readRDS("test/sp17-enrolled.rds")
target_summary = readRDS("test/summary-sp17.rds")

# Verify existence of objects
expect_true(exists("overview_sp17"), info = "Verify summary was created.")
expect_true(exists("sp17_enrolled"), info = "Verify graph was created.")

# Verify data is accurate
expect_equal(overview_sp17, target_summary, info = "Verify summary was obtained correctly.")
expect_equal(sp17_enrolled, target_df, info = "Verify data frame was created correctly.")
```



## (17 Points) Exercise 2: The Recycler

**To receive credit for answers to exercises listed here, you _must_ 
use _R_'s recycling behavior.**

_Hint_ You may wish to consult help documentation for the `rep()` function and
revisit the vectorization slides at the 
[Transforming Data](http://stat385.thecoatlessprofessor.com/lectures/05-transforming-data/05-transforming-data.pdf#page=32)
and accompanying 
[code](https://dl.dropboxusercontent.com/s/b8cdo82eaib6uul/09-07-2018-transforming-data.Rmd?dl=0).

- **[2 Points] (a)** If we define `x = c(2, 10)` and `y = c(9, 2, 6, 10)`,
   what should the output of `z = x * y` be? 
   

```{r vectorized-test, test = TRUE}
target_vec = readRDS("test/z-result.rds")
expect_equal(z, target_vec)
```


- **[3 Points] (b)** Explain why the output is occuring by:
    - Creating an example using `rep()` that mimics
      how _R_ is performing the operation.
    - Describe the underlying property _R_ applies to vectors.


- **[6 Points] (c)** What happens if we update `x` to be
    `x = c(3, 5, 12, 18, 11)` and perform the operation again? 
    - **[1 Points]** Recompute the result of `x*y` with the new `x`
    - **[4 Points]** Create and include a _diagram_ that shows the 
     process _R_ goes through when evaluating this operation.
        - The included diagram can be a picture of a _hand_ sketch, digitally drawn(MS Paint, PowerPoint, Photoshop, ... ), or even done using [`DiagrammeR`](http://rich-iannone.github.io/DiagrammeR/).
        - The choice is yours.




- **[3 Points] (d)** Create the `alt_vec` vector with length 31 such that 
      successive values alternating between polarity (e.g. `1` and `-1`). 
    - For example, the first _four_ values of such a vector would be: 
      `c(1, -1, 1, -1)`
    - _Hint:_ Consider $(1)^2$, $(-1)^2$, $(1)^3$, and $(-1)^3$.
    - _Hint:_ The colon (`:`) operator is useful to generate sequences.



```{r seq-gen-alt-issue, test = TRUE}
target_vec = readRDS("test/alternating-vec.rds")

expect_equal(alt_vec, target_vec)
```

- **[3 Points] (e)** Construct the following sequence `k` of length 15:
   $1, -2, 3, 4, -5, 6, 7, -8, 9, \ldots, 13, -14, 15$.
   - This sequence _must_ be created using recycling...




```{r seq-gen-custom-issue, test = TRUE}
target_vec = readRDS("test/reconstructed-seq.rds")

# expect_equal(k, target_vec)
```

## (8 Points) Exercise 3: Scientists Rock!

**[3 Points] (a)** Develop a function called `fahrenheit_to_celcius()` that converts
   fahrenheit (째F) to celcius (째C) given the definition of: 
   $$f(x) = \frac{{5\left( {x - 32} \right)}}{9}$$

```{r my-fahrenheit}
#' @title Convert from Fahrenheit to Celcius
#' @description
#' Given a temperature in Fahrenheit convert it to the
#' the temperature in Celcius.
#'
#' @param degrees_c A `numeric` vector containing temperature in degrees Celcius.
#'
#' @return A `numeric` vector of temperatures in degrees Fahrenheit.
#' @details
#' The function implements the following formula: 
#' \eqn{f(x) = \frac{{5\left( {x - 32} \right)}}{9}}{5*(x-32)/9}
#'
#' @examples
#' # Calculate a single value
#' fahrenheit_to_celcius(32.0)
#' 
#' # Calculate multiple values
#' fahrenheit_to_celcius(c(69.8, 98.6, 212.0))
fahrenheit_to_celcius = function(degrees_c) {
 
  # Logic implemented here
   
}
```


```{r fahrenheit-test, test = TRUE}
# Check single value
expect_equal(
  fahrenheit_to_celcius(32),
  0, 
  info = "Check single fahrenheit value"
)

# Check vectorization
expect_equal(
  fahrenheit_to_celcius(c(69.8, 98.6, 212.0)),
  c(21, 37, 100), 
  info = "Check multiple fahrenheit values"
)
```


**[5 Points] (b)** Now, create a function called `celcius_to_fahrenheit()` that
does the inverse of the previous one. That is, write $f^{-1}(x)$ which
translates celcius (째C) to fahrenheit (째F).

- **[2 Points]** Derive the formula for the celcius to fahrenheit conversion
  and include it in the `@details` section of the function's documentation.
- **[3 Points]** Implement the derived formula in the _R_ function below.

```{r my-celcius}
#' @title Convert from Celcius to Fahrenheit
#' @description
#' Given a temperature in Celcius convert it to the
#' the temperature in Fahrenheit.
#'
#' @param degrees_f A `numeric` vector containing temperature in degrees Fahrenheit.
#'
#' @return A `numeric` vector of temperatures in degrees Celcius.
#' @details
#' The function implements the following formula: 
#' \eqn{LATEX-HERE}{REGULAR-SYMBOLS-HERE}
#' An example of this is given above...
#' @examples
#' # Calculate with a single value
#' celcius_to_fahrenheit(0)
#' 
#' # Calculate multiple values
#' celcius_to_fahrenheit(c(21, 37, 100))
celcius_to_fahrenheit = function(degrees_f) {
 
  # Logic implemented here
   
}
```


```{r celcius-test, test = TRUE}
# Check single value
expect_equal(
  celcius_to_fahrenheit(c(0)),
  32, 
  info = "Check single celcius value"
)

# Check vectorization
expect_equal(
  celcius_to_fahrenheit(c(21, 37, 100)),
  c(69.8, 98.6, 212.0), 
  info = "Check multiple celcius values"
)
```

## (10 Points) Exercise 4: Winter is Coming

**[2 Points] (a)** Install and load the `nasaweather` package. 
**Comment** out the installation command in your `.Rmd` file. 
(If you do not comment installation commands out, then they will be run
every time you knit your `.Rmd` file.) 



**[2 Points] (b)** Open up the help documentation for `glaciers` (the data set),
find where the variables for the data set are listed and write in your
_RMarkdown_ document _what_ the `country`  variable contains.


**[6 Points] (c)** Provide summary information on the `glaciers` by:

- writing a sentence that _dynamically_ describes the dimensions of the data;
    - A _dynamic_ description means that sentences should use inline _R_ code.
      Examples can be found in [Literate Programming]()
- showing the last ten observations in the data set; and,
- providing a summary overview of the data to understand its contents.





##  (14 Points) Exercise 5: Excellency at UIUC

Under this exercise, we will explore the "Teachers Ranked As Excellent" data
at UIUC from Fall 1993 to Spring 2018 as compiled by 
[Wade Fagen-Ulmschneider](http://waf.cs.illinois.edu/).
Please download the data from:

<https://raw.githubusercontent.com/wadefagen/datasets/master/teachers-ranked-as-excellent/uiuc-tre-dataset.csv> 

To download the data, go to the webpage and press `Cntrl/Cmd + S` to save the
file to your local computer. Name the file `uiuc-tre-dataset.csv`. _Upload it_
onto RStudio Cloud. For help, see [**Page 20** of **Reading 0**](https://github.com/stat385uiuc/rstudio-cloud-intro/raw/master/01-rstudio-cloud-stat385.pdf).

This data has a file extension of **CSV** form. Contained in the data are the
following variables:

- `term`: Two letter semester code (`sp`, `su`, `fa`, or `wi`) followed by a four digit year.
    - Examples: `sp2017`, `fa2013`, `su2012`.
- `unit`: The CITL-supplied headers for the unit teaching the course.
    - Examples: `ACCOUNTANCY`, `SOCIAL WORK`, `LINGUISTICS`, `NUCLEAR, PLASMA & RAD. ENGR.`
- `lname`: The last name of the teacher.
    - Examples: `FAGEN-ULMSCHNEIDER`, `FLANAGAN`, `FLECK`
- `fname`: The first letter of the first name of the teacher.
    - Examples: `W`, `K`, `M`
- `role`: `Instructor` or `TA`
- `ranking`: `Excellent` or `Outstanding`
- `course`: The course the teacher was ranked as excellent.  If no course is given, the `course` field is set to `?` (this includes cases when the raw data lists the course as `0`, `000`, or `999`).
    - Examples: `199`, `225`, `560`, `?`

**[2 points] (a)** Import into _R_ the data in `uiuc-tre-dataset.csv` with the 
variable name `teaching_list_data`. As `course` contains a value that is
_not_ `NA`, which is how _R_ represents missing values, you must use the parameter 
`na.strings = c("?", "NA")` in the appropriate `read.*()` function. 
If you need help, please see the appropriate help documentation via `?`.

**NB** The `*` in the `read.*()` statement is a placeholder for the type of file
you want to read in. 


**[2 points] (b)**  Retrieve the dimensional information for this data using
only one function.


**[4 points] (c)** Perform a summary of the `teaching_list_data`. Within the summary
output, what variable output is different from the rest? What might have caused this? 
_Hint_ Consider looking at the structure of `teaching_list_data`. 




**[2 points] (d)** Who was your favorite teacher at UIUC? Search their
last name in all capitals.

```{r my-fav-teacher, eval = FALSE}
teaching_list_data[teaching_list_data$lname == "YOUR TEACHERS LAST NAME HERE IN CAPITALS", ]
```



**[4 points] (e)** Fix the following code so that it:

1. doesn't error;
1. produces a graph; and 
1. hides the code.

_Hint_ for the last two requirements look at different code chunk options.

```{r ggplot2-code, eval = FALSE}
ggplot(teaching_list_data, aes(fname)) + 
  geom_bar() +
  facet_wrap( ~role) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Frequency",
       x = "First Name Letter")
```


